# (PART) Introduction {-}

# Introduction to R {#lab1}

```{r include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  strip.white = TRUE,
  comment = "#>",
  out.width = "65%",
  message=FALSE,
  warnings=FALSE
)
```

```{r, include=FALSE}
tutorial::go_interactive()
```

Statistical analysis is the core of nearly all research projects and researchers have a wide variety of statistical tools that they can use, like *SPSS* and *SAS*. Unfortunately, these analysis tools are expensive or difficult to master so this lab manual introduces R, a powerful, open source statistical analysis program that is available free of charge. But before diving into a statistics package there is one important background fundamental that must be covered: data.

## Types of Data

There are two main types of data and it is important to understand the difference between them since that determines appropriate statistical tests to be used in data analysis.

* Continuous data are integer or decimal numbers and are typically used for counts or measures -- like a person's weight, a tree's height, or a car's speed. Continuous data are measured with scales that have equal divisions so the difference between any two values can be calculated. Because continuous data include characteristics like means and standard deviations, they are analyzed using parametric tests, as described in [Lab 9](#lab9).

* Categorical data group observations into a limited number of categories; for example, type of pet (cat, dog, bird, etc.) or place of residence (Arizona, California, etc.). One common type of categorical data is generated with an "agree-disagree" type of scale, like "I enjoy reading: Strongly Agree -- Agree -- Neutral -- Disagree -- Strongly Disagree." Because categorical data do not have characteristics like means or standard deviations, they are analyzed using nonparametric tests, as described in [Lab 10](#lab10).

## Fundamental R Operations

### Command Line Operations

All R commands are entered from a "Command Line" environment. Many students find this a bit challenging at first but once they learn some foundational concepts the command line becomes easy and fast to use. This is an explanation for the R script in the box below.

* Line 1: This is a comment that is used to record notes in a script. In R, all comments start with a hash-mark (#) and everything after that symbol is ignored. Comments are used frequently in scripts presented in this manual in order to explain what the script is doing. Good programmers comment liberally so when a team member has to figure out what they did it is easy to understand.

* Line 2: Calculate the value of 3+5.

* Line 3: Calculate the value of "5 + 8 * 2".

* Lines 6-7: These lines create two variables, *MaxScore* and *MinScore*, and then assign values to the variables. You should note two important things about these lines. First, the "assignment" operator is a less than sign followed by a hyphen, like <-, making a leftpointing arrow. That tells R to store the number on the right side of the arrow operator into the variable named on the left side of the line. Also, keep in mind that capitalization matters with R. Thus, the variable named *MaxScore* would be different than a variable named *maxscore*. These lines only store values in variables and nothing gets printed to the screen.

<div class="grsnote">
A *variable* is nothing more than a place in memory to store temporary data. Think of it as a "box" that is used to store something until it is needed later.
</div>

* Line 8: The variable *Range* is filled with the result of subtracting *MinScore* from *MaxScore*.

* Line 9: Entering a variable name, like *Range*, on a line by itself causes the value stored in that variable to be displayed.

* Line 12: In R, a list of numbers can be stored in a single variable by using the "combine" function, which is a *c* followed by a list of the numbers inside a parenthesis. This line creates a variable called *TestScores* and then stores a list of six numbers in that variable.

* Line 13: The contents of the variable *TestScores* is printed to the screen.

```{r}
# Some basic calculations
3 + 5
5 + 8 * 2

# Using variables
MaxScore <- 57
MinScore <- 22
Range <- MaxScore - MinScore
Range

# The combination function
TestScores <- c(85, 83, 92, 88, 71, 85)
TestScores
```

<div class="grsnote">
The *DataCamp* blocks found throughout this lab manual are designed to "try out" R commands. Click the yellow "Run" button in the lower left corner to execute the entire script and see the results. There are two panels available in the box. Clicking *script.R* in the top of the box displays the R script and clicking *R Console* displays the results of the script in an R console. The commands in the script can be modified and rerun in order to "play around" with R. Also, R commands can be entered directly in the Console and executed by tapping the [ENTER] key.
</div>

### Data Frames

A *data frame* is a collection of data generated during a research project. An example data frame that is easy to understand would be a spreadsheet that contains the times recorded for a race. R comes configured with a 103 built-in data frames used for training and the R script below is an introduction to one of the data frames used in several of the labs in this manual: *mtcars*. (*Note*: [Appendix A](#datadic) has more information about the data frames used in this lab manual.) 

* Line 2: Entering the name of the data frame, *mtcars*, on a line by itself causes R to print the contents of the entire the data frame to the screen. Since *mtcars* is rather small it is fine to print it to the screen, but larger data frames may cause the screen to "scroll" for some time before the end of the data frame is reached.

* Line 3: This prints the structure of the *mtcars* data frame. The result shows that this is a *data.frame* and has 32 observations (that is how many cars are in the dataset) of 11 variables (things like *mpg*). Also the structure command displays the type of data that are in the dataset. For example, all 11 variables are of the "number" type. The *str* function is frequently used to better understand a data frame.

* Line 4: This line prints the maximum *mpg* value for the *mtcars* data frame. Note that the specific variable desired is indicated by both the data frame name and the variable, separated by a dollar sign, like *mtcars$mpg* on this line.

* Line 5: The attach command is used to instruct R that all following commands should be applied to the attached data frame. This makes it easier to enter commands since the data frame name does not need to be specified for every command.

* Line 6: This prints the minimum *mpg* value. Notice that the data frame name does not need to be specified since *mtcars* was attached on Line 5.

*The result for some of the lines in this script are too long to fit on one row in the Console so they wrap around.*


```{r}
# The Motor Trend Cars Data Frame
mtcars
str(mtcars)
max(mtcars$mpg)
attach(mtcars)
min(mpg)
```

### Main Street Cafe Data

The *Main Street Cafe* data frame will be used for all of the submitted activities in this manual and this lab is an introduction to that data frame. This is an explanation for the R script in the box below.

* Line 1: This line displays the structure of the cafe data frame. *Cafe* contains 375 observations (that is, 375 cafe customers were surveyed) of 13 variables (those 375 customers were asked 13 questions). Note that some of the variables are factors, some integers, and one is a number. Factors are ordinal or nominal variables with "levels" of response, like "female," "male," and "Other." Numeric data are divided into two types: Integers have no decimal part and "num" include a decimal part.

* Line 2: The *cafe* data frame is attached to the script so later commands will automatically access it.
* Line 3: The maximum value of the "age" variable from the *cafe* data frame is displayed.
* Line 4: The minimum value of the "age" variable from the *cafe* data frame is displayed.

```{r ex="cafeData1", type="pre-exercise-code"}
cafe <- read.csv('http://georgeself.com/maincafe.csv')
```

```{r ex="cafeData1", type="sample-code"}
str(cafe)
attach(cafe)
max(age)
min(age)
```

## Activities

### Activity 1

Start R and calculate the following problems:

* 5 + 8 * 3
* 10 * 2/4
* 5 * 3 * 3 - 20

Include the answers to these three problems in the deliverable document for this lab.

```{r}
# Calculate the problems listed above.


```


### Activity 2

Look at the structure for the *rock* dataset. Note the number of observations and variables present in that dataset. Include the answers to these questions in the deliverable document for this lab.

```{r}
# Look at the structure for the rock dataset and note the number of observations and variables.


```

### Activity 3

Using the *cafe* data, find the maximum and minimum values for the *bill* variable and include those values in the deliverable document for this lab. The structure of the cafe data frame does not need to be submitted for this activity, only the maximum and minimum values for the "bill" variable. *Note: remember to either attach cafe or include that data frame name in the min and max functions or an "object not found" error will pop up.*

```{r ex="act1.3", type="pre-exercise-code"}
cafe <- read.csv('http://georgeself.com/maincafe.csv')
```

```{r ex="act1.3", type="sample-code"}
# Attach the cafe data and then find the maximum and minimum values of the bill variable.


```


## Deliverable

Complete the activities above and consolidate the responses into a single document. Name the document with your name and "Lab 1," like "George Self Lab 1" and submit that document for grade. It is also acceptable to consolidate the responses into a Google Document and submit a link to that document.
